---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r}

library(lattice)

# load the data
activity<-read.csv("activity.csv")

# process the data
# convert date column to date type
activity$date<-as.Date(activity$date)

# only use observations without NA values
clean_activity<-activity[!is.na(activity$steps),]

```


## What is mean total number of steps taken per day?

1. Calculate the total number of steps taken per day
```{r}

# calculate the sum of steps per day
part1<-aggregate(steps ~date, clean_activity, sum)
```
2. Make a histogram of the total number of steps per day
```{r, echo=TRUE}
# plot the histogram (this one will be blue)
hist(part1$steps, 
     main= "Total number of steps per day",
     xlab="Steps", 
     breaks=15,
     col="blue")
```


3. Calculate and report mean and median of the total number of steps take per day
```{r, echo=TRUE}
#calculate mean and median
mean(part1$steps)
median(part1$steps)
```
The mean and median are very close

## What is the average daily activity pattern?
1. Make a time series plot of the average number of steps taken during a time interval
```{r, echo=TRUE}

# calculate average steps per interval
interval_steps<-aggregate(steps ~ interval, clean_activity, mean)

# plot average steps per interval
plot(interval_steps$interval,
     interval_steps$steps, 
     type="l",
     main="Average number of steps taken during a given time interval",
     xlab="Time interval", 
     ylab="Number of steps")
```

2. Which interval contains the maximum number of steps? 
```{r, echo=TRUE}

# Use which.max() to find largest mean number of steps per day,
# then return the associated interval
interval_steps[which.max(interval_steps$steps), 1]


```
The interval with the hightest mean number of steps is 835.

## Imputing missing values
1. Calculate and report the total number of missing values in the dataset
```{r, echo=TRUE}
# count the number of NA values in the orignal dataset
nrow(activity[is.na(activity$steps),])
```

The total number of missing values in the dataset is 2304.

2. Devise a strategy for filling in all missing values in the dataset.

The strategy is to replace "NA" values with the average steps taken for
    that time interval across all days
  
3. Creae a new dataset that is equal to the original dataset but with NAs filled in. 
```{r}

# copy original dataset
fixed_activity<-activity

# replace NA values with average number of steps for that time interval
fixed_activity$steps[is.na(fixed_activity$steps)]<-interval_steps$steps

# calculate total number steps per day of imputed data 
part3<-aggregate(steps ~date, fixed_activity, sum)
```


4a. Make a histogram of the total number of steps taken in the imputed dataset.

```{r, echo=TRUE}

# plot a histogram (this one will be green)
hist(part3$steps, 
     main= "Total number of steps per day (Missing values imputed)",
     xlab="Steps", 
     breaks=15,
     col="green")
```

4b. Calculate and report the mean and median total number of steps taken per day of the imputed
data set
```{r, echo=TRUE}
#calculate mean and median
mean(part3$steps)
median(part3$steps)
```

The mean is unchanged and the median is now slightly larged (and the same value as the mean).
It does not appear that imputing the missing values had much of an effect on the total daily
estimated number of steps.

## Are there differences in activity patterns between weekdays and weekends?

1. Create a new factor variable in the dataset: weekday or weekend.
```{r}

# use weekdays() function to label each day. 
# Notice, the dataset without NAs is used
clean_activity$weekend <- ifelse(weekdays(clean_activity$date) %in% 
                c("Saturday", "Sunday"),
                "weekend", "weekday")
clean_activity$weekend<-as.factor(clean_activity$weekend)

# same strategy as before but add the weekend factor variable
avg_steps_w<-aggregate(steps ~ interval + weekend, clean_activity, mean)
```

2. Make a panel plot of the average number of steps taken during an interval (on the weekend or during the week)

```{r, echo=TRUE}

#lattice plot separated by weekend and weekday
xyplot(avg_steps_w$steps ~ avg_steps_w$interval | avg_steps_w$weekend, 
       type="l", 
       layout=c(1,2),
       main="Comparison between activity on weekdays and weekends",
       xlab="Time intervals",
       ylab="Average steps")
```

There appears to be similar activity during the week in the morning, however there is more activity during the day on the weekend (perhaps the subject is at work at a desk during the week?)